Thermal reactionomes reveal adaptive responses to thermal

### 2014-10-08 doc/ApTranscriptome_manuscript.Rmd

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@gilchrist1995; @angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to test two specific predictions. First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance (e.g. "frontloading", [@barshis2013]) than plastically-induced resistance mechanisms relative to species from cooler environments. This pattern could be manifested through the number of thermally responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Specifically, finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased constitutive expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-26.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

@@

## Transcripts differ in thermal responsiveness between species

@@

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
```

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).

We found that the number of transcripts that responded to low temperatures is substantially greater than of that to high temperatures (Table 1). In both species, the majority of transcripts were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate**.

Gene set enrichment analysis (GSEA) of the responsive transcripts indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** transcripts. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *High* transcripts, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 8 of the 9 terms being involved in regulation of biosynthetic or metabolic process".  For **Bimodal** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "transcript expression", "signaling", and "protein modification process". For **Intermediate** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".

# Shifts in the expression patterns of thermally responsive transcripts between species

Of the thermally responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. This was not due to a species difference in the overall number of responsive genes (...stat...). Instead, most of these were transcripts that changed from one expression category to another (i.e. the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...) indicating that these differences are not simply due to stochastic processes.

**Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

@@

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}
#tt2

# marginal sums and total
rs <- rowSums(tt2)
cs <- colSums(tt2)
n <- sum(tt2)

# grand mean for each class
gm <- (rs + cs) / sum(tt2 * 2)

# calculate observed values for each cell using overall mean for each expression type
Ec <- outer(gm, gm, "*") *sum(tt2)

# get deviations of observed from expected
Ec.dev <- tt2 - Ec

# overall table P-value
Ec.chistat <- sum((tt2-Ec)^2 / Ec)
# P-value
pchisq(Ec.chistat, df=(nrow(tt2)-1) * (ncol(tt2)-1), lower.tail = FALSE)

# overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
df = (ncol(tt2) - 1) * (nrow(tt2) - 1)
X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)

# posthoc result for each cell
Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
                                                            ifelse(x > X2.df, "+", "-")))

md <- melt(Ec.cells)

@@

## Species differences in the sensitivity of thermally responsive transcripts to changes in temperature

To examine how the sensitivity of transcript expression to changes in temperature differed between the two species, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation. Transcripts turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Consistent with the results for up-regulated transcripts, we found greater inertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea* drive by a differential response to heat stress. With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

## Adaptive changes in constitutive expression are not linked to transcripts involved in short-term physiological acclimation

We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons. First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression. Thus, the warm-adaptated *A. carolinensis* should display higher constitutive expression of **High** transcripts than *A. picea*, while the reverse should be true for **Low** transcripts.  Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for transcripts that had **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

In general, thermally responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

The only exception was the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.

## Greater up-regulation of heat shock proteins in the cool-climate species *A. picea*

We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally responsive transcripts is consistent with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

@@

A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2013]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species. Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation. Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].

@@

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions (.....) .  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

## Thermally responsive transcripts

@@

We identified thermally responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

For each thermally responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression values to partition the thermally responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

>>>>

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test a number of alternative mechanistic hypotheses to explain how populations evolutionarily increase their thermal breadth in response to stressful environments. First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species recruits additional molecular pathways to provide greater protection from damage. This hypothesis predicts that *A. picea* induces more genes in response to cold temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. An additional prediction of this hypothesis is that upregulation of the shared molecular pathways should be strong or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts that the two species should differ in the identity and timing of gene induction in response to temperature changes. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-10-08.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

@@

## Thermally responsive gene responsed non-linearly to changes in temperature

@@

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
```

## *A. picea* recruits additional molecular pathways in response to heat stress

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The expression pattern differed between species for `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) of the thermally-responsive transcripts.

Consistent with the *enhanced response* hypothesis, *A. picea* induced expression of about 1,000 (30%) more transcripts than *A. carolinensis* in response to cold temperatures (Table 1, ...stats...). However, *A. picea* also induced expression of more transcripts, though to a smaller extent (~14%), than did *A. carolinensis* in response to high temperatures and induced more transcripts that are responsive to both high low temperatures (*Bimodal* transcripts; Table 1).

The additional genes recruited included molecular processes such as...GO terms...

## *A. carolinensis* shuts-down gene expression in response to thermal extremes

While *A. picea* induced more genes, the total number of thermally-responsive transcripts did not differ between species (...stat...). Instead, most of these transcripts shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

@@

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}

@@

## Greater constitutive expression of thermally-responsive genes in *A. carolinensis*

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by comparing constitutive expression, here defined as expression at 19.5°C, between species for high temperature responsive transcripts in *A. picea* and *A. carolinensis*. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for high temperature responsive transcripts in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`). However, this pattern was also true for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`). Thus, the overall pattern is that thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

An exception to this overall pattern is the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*.

## *A. carolinensis* tolerates heat stress to a greater extent than *A. picea*

To test the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes, and the identity of these genes. Specifically, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation. In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Again, consistent with the *tolerance* hypothesis, we found greater inertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea* drive by a differential response to heat stress. With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

@@

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is consistent with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

**put heat shock protein in Discussion**

> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species. Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation. Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].

@@

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

## Thermally-responsive transcripts

@@

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
